Contracts for Haskell

### 2011-06-30 contracts.tex

\begin{center}
\begin{array}{rclr}
  u,e &:=& x \mid f \mid \lambda x.~e \mid e~e \mid D(e,\dots,e) \mid \bad \mid n \mid e \mbox{ `satisfies` } c \mid \mbox{assert } (e,c) \mbox{ in } e  & (Expression)\\
  &\mid& \mbox{case } e \mbox{ of } [(p_i,e_i)] &\\
  p &:=& \Delta,p \mid T,p \mid f \in c,p \mid \epsilon & (Program)\\
  \Delta &:=& d \mid opaque(d) & (Defintion~and~scope)\\
  d &:=& f~x_1 \dots x_n ::: c = e \mid f~x_1 \dots x_n ::: c = \mbox{ case } e \mbox{ of } [(pat_i,e_i)] & (Definition)\\
  T &:=& \mbox{data } x = DCons  & (Data~type~definition)\\ %TODO make type clearer
  DCons &:=& \epsilon \mid D ; DCons \mid D ::: c ; DCons & (List~of~data~constructors)\\
  pat &:=& D(x_1,\dots,x_n) & (Pattern)\\
\end{array}
\end{center}

@@

\begin{center}
\begin{array}{rclr}
  u,e &:=& x \mid f \mid e~e \mid D(e,\dots,e) \mid \bad \mid n \mid e \mbox{ `satisfies` } c & (Expression)\\
  p &:=& \Delta,p \mid T,p \mid f \in c,p \mid \epsilon & (Program)\\
  \Delta &:=& d \mid opaque(d) & (Defintion~and~scope)\\
  d &:=& f~x_1 \dots x_n ::: c = e\mid f~x_1 \dots x_n ::: c = \mbox{ case } e \mbox{ of } [(pat_i,e_i)] & (Definition)\\
  T &:=& \mbox{data } x = DCons  & (Data~type~definition)\\ %TODO make type clearer
  DCons &:=& \epsilon \mid D ; DCons \mid D ::: c ; DCons & (List~of~data~constructors)\\
  pat &:=& D(x_1,\dots,x_n) & (Pattern)\\
\end{array}
\end{center}

@@

\subsection{Contracts}
\begin{center}
\begin{array}{rclr}
  c &:=& x:c_1 \to c_2\\
  &\mid& (c_1,c_2) \\
  &\mid& \{ x \mid e \}\\
  &\mid& \{ !x \mid e \}\\
\end{array}
\end{center}

Semantics of contract satisfaction (exactly the same as in the POPL paper):
\begin{center}
\begin{array}{rcl}
  e \in \{x \mid p \} &\iff& e \mbox{ diverges or } p[e/x] \not \to^\star \{\bad,False\}\\
  e \in \{!x \mid p \} &\iff& e \mbox{ diverges or } (e \mbox{ is crash-free and } p[e/x] \not \to^\star \{\bad,False\})\\
  e \in x:t_1 \to t_2 &\iff& \forall e_1 \in t_1, (e~e_1) \in t_2[e_1/x]\\
  e \in (t_1,t_2) &\iff& e \mbox{ diverges or } (e \to^\star (e_1,e_2) \mbox{ and } e_1 \in t_1, e_2 \in t_2)\\
\end{array}
\end{center}

@@

\begin{eqnarray}
\etrans{x} &=& (x,\emptyset)\\
\etrans{f} &=& (f,\emptyset)\\
\etrans{e_1~e_2} &=& (app(\etranst{e_1},\etranst{e_2}),\etransf{e_1} \cup \etransf{e_2})\\
\etrans{D(e_1,\dots,e_n)} &=& (D(\etranst{e_1},\dots,\etranst{e_n}),\cup_{1 \leq i \leq n} \etransf{e_i})\\
\etrans{\bad} &=& (\bad,\emptyset)\\
\etrans{n} &=& (n,\emptyset)\\
\etrans{e \mbox{ `satisfies` } c} &=& (\etranst{e},\etransf{e} \cup \forall x. \strans{x \in c} \leftrightarrow x~`satisfies`~c = True)\\
\end{eqnarray}

@@

\subsection{$\strans{}$}
$\strans{e \in c}$ is a set of first-order formulae.
\begin{eqnarray}
\strans{e \in \{x \mid u \}} &=& e = \unr \lor (\mbox{CF}(\etranst{e}) \land \etranst{u[e/x]} \neq \bad \land \etranst{u[e/x]} \neq False)\\
&& \cup \etransf{e} \cup \etransf{u[e/x]}\\
\strans{e \in x:c_1 \to c_2}  &=& \forall x_1. \strans{x_1 \in c_1} \to \strans{e~x_1 \in c_2[x_1/x]}
\end{eqnarray}
$False$ is a data constructor here.

>>>>

\begin{center}
\begin{array}{rclr}
  u,e &:=& x \mid f \mid \lambda x.~e \mid e~e \mid D(e,\dots,e) \mid \bad \mid n \mid e \mbox{ `satisfies` } c \mid \mbox{assert } (e,c) \mbox{ in } e  & (Expression)\\
  &\mid& \mbox{case } e \mbox{ of } [(p_i,e_i)] \mid \cf{e}&\\
  p &:=& \Delta,p \mid T,p \mid f \in c,p \mid \epsilon & (Program)\\
  \Delta &:=& d \mid opaque(d) & (Defintion~and~scope)\\
  d &:=& f~x_1 \dots x_n ::: c = e \mid f~x_1 \dots x_n ::: c = \mbox{ case } e \mbox{ of } [(pat_i,e_i)] & (Definition)\\
  T &:=& \mbox{data } x = DCons  & (Data~type~definition)\\ %TODO make type clearer
  DCons &:=& \epsilon \mid D ; DCons \mid D ::: c ; DCons & (List~of~data~constructors)\\
  pat &:=& D(x_1,\dots,x_n) & (Pattern)\\
\end{array}
\end{center}

@@

\begin{center}
\begin{array}{rclr}
  u,e &:=& x \mid f \mid e~e \mid D(e,\dots,e) \mid \bad \mid n \mid e \mbox{ `satisfies` } c \mid \cf{e} & (Expression)\\
  p &:=& \Delta,p \mid T,p \mid f \in c,p \mid \epsilon & (Program)\\
  \Delta &:=& d \mid opaque(d) & (Defintion~and~scope)\\
  d &:=& f~x_1 \dots x_n ::: c = e\mid f~x_1 \dots x_n ::: c = \mbox{ case } e \mbox{ of } [(pat_i,e_i)] & (Definition)\\
  T &:=& \mbox{data } x = DCons  & (Data~type~definition)\\ %TODO make type clearer
  DCons &:=& \epsilon \mid D ; DCons \mid D ::: c ; DCons & (List~of~data~constructors)\\
  pat &:=& D(x_1,\dots,x_n) & (Pattern)\\
\end{array}
\end{center}

@@

\subsection{Contracts}
\begin{center}
\begin{array}{rclr}
  c &:=& x:c_1 \to c_2\\
  &\mid& (c_1,c_2) \\
  &\mid& \{ x \mid e \}\\
  &\mid& c_1 \land c_2 \\
  &\mid& c_1 \lor c_2 \\
\end{array}
\end{center}

Semantics of contract satisfaction (exactly the same as in the POPL paper):
\begin{center}
\begin{array}{rcl}
  e \in \{x \mid p \} &\iff& e \mbox{ diverges or } p[e/x] \not \to^\star \{\bad,False\}\\
  e \in x:t_1 \to t_2 &\iff& \forall e_1 \in t_1, (e~e_1) \in t_2[e_1/x]\\
  e \in (t_1,t_2) &\iff& e \mbox{ diverges or } (e \to^\star (e_1,e_2) \mbox{ and } e_1 \in t_1, e_2 \in t_2)\\
  e \in c_1 \land c_2 &\iff& e \in c_1 \mbox{ and } e \in c_2\\
  e \in c_1 \lor c_2 &\iff& e \in c_1 \mbox{ or } e \in c_2\\
\end{array}
\end{center}

@@

\begin{eqnarray}
\etrans{x} &=& (x,\emptyset)\\
\etrans{f} &=& (f,\emptyset)\\
\etrans{e_1~e_2} &=& (app(\etranst{e_1},\etranst{e_2}),\etransf{e_1} \cup \etransf{e_2})\\
\etrans{D(e_1,\dots,e_n)} &=& (D(\etranst{e_1},\dots,\etranst{e_n}),\cup_{1 \leq i \leq n} \etransf{e_i})\\
\etrans{\bad} &=& (\bad,\emptyset)\\
\etrans{n} &=& (n,\emptyset)\\
\etrans{e \mbox{ `satisfies` } c} &=& (\etranst{e},\etransf{e} \cup \forall x. \strans{x \in c} \leftrightarrow x~`satisfies`~c = True)\\
\etrans{\cf{e}} &=& \cf{e}\\
\end{eqnarray}

@@

\subsection{$\strans{}$}
$\strans{e \in c}$ is a set of first-order formulae.
\begin{eqnarray}
\strans{e \in \{x \mid u \}} &=& e = \unr \lor (\etranst{u[e/x]} \neq \bad \land \etranst{u[e/x]} \neq False)\\
&& \cup \etransf{e} \cup \etransf{u[e/x]}\\
\strans{e \in x:c_1 \to c_2}  &=& \forall x_1. \strans{x_1 \in c_1} \to \strans{e~x_1 \in c_2[x_1/x]}\\
\strans{e \in c_1 \land c_2} &=& \strans{e \in c_1} \land \strans{e \in c_2}\\
\strans{e \in c_1 \lor c_2} &=& \strans{e \in c_1} \lor \strans{e \in c_2}\\
\end{eqnarray}
$False$ is a data constructor here.
